GEE models

### 2017-10-02 doc/manuscript.Rmd
data availability statement
beta-cell -> beta cell

Title: Association of non-esterified fatty acid composition with insulin
sensitivity and beta-cell dysfunction in the Prospective Metabolism and Islet Cell
Evaluation (PROMISE) cohort

@@

Aim: Our aim was to determine the longitudinal associations of individual non-esterified fatty acids (NEFA) on differences in insulin sensitivity (IS) and beta-cell function over 6 years in a cohort of individuals who are at-risk for diabetes.

Methods: In the PROMISE cohort, 477 participants had serum NEFA (mol% and concentration) measured at the baseline visit and completed an OGTT at 3 time points over 6 years. Outcome variables were calculated using the OGTT values. IS was assessed using HOMA2-%S and the Matsuda index, while beta-cell function was  assessed using the Insulinogenic index over HOMA-IR (IGI/IR) and the Insulin  Secretion-Sensitivity Index-2 (ISSI-2). Generalized estimating equations were  used, adjusting for time, waist, sex, ethnicity, baseline age, alanine  aminotransferase, and physical activity. NEFA were analyzed as both  concentrations (nmol/mL) and proportions (mol%) of the total fraction.

Results: Participants (73% female, 70% with European-ancestry) had their IS and beta-cell function decline by 14-27% over 6 years of follow-up. While in unadjusted models there were several NEFA (e.g. 18:1n-7, 20:1n-9, 22:4n-6) that had negative associations with insulin sensitivity, nearly all of these associations were attenuated in fully adjusted models. In adjusted models, total NEFA, 16:0, 18:1n-9, and 18:2n-6 (as concentrations) had significant negative associations with IGI/IR and ISSI-2, while only 20:5n-3 (as mol%) had positive associations with HOMA2-%S.

Conclusions: Total NEFA concentration was a strong predictor of lower beta-cell  function over 6 years. Our results suggest that the association with beta-cell  function is due to the absolute size of the serum NEFA fraction, rather than the specific fatty acid composition.

@@

insulin sensitivity; beta-cell function; longitudinal cohort; non-esterified
fatty acids; diabetes pathogenesis; fatty acid composition

@@

tab_basic <- tab('basic', 'Basic characteristics of the PROMISE participants at each of the three clinic visits.')
fig_dist <- fig('nefa', 'Concentrations (nmol/mL) of each NEFA in PROMISE participants at the baseline visit (2004-2006).')
fig_gee_unadj <- fig('gee_unadj', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function over 6 years in the PROMISE cohort. Models are only adjusted for time. X-axis values represent a percent difference in the outcome per SD increase in the fatty acid. P-values were adjusted for the BH false discovery rate, with each increase in dot size and blackness representing a p-value significance of p>0.05, p<0.05, p<0.01, and p<0.001.')
fig_gee_adj <- fig('gee_adj', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function over the 6 years in the PROMISE cohort. Generalized estimating equation models were fully adjusted. X-axis values represent a percent difference in the outcome per SD increase in the fatty acid. P-values were adjusted for the BH false discovery rate, with each increase in dot size and blackness representing a p-value significance of p>0.05, p<0.05, and p<0.01.')
```

@@

Total non-esterified fatty acids (NEFA) have been well documented to influence 
the pathogenesis of type 2 diabetes mellitus. Experimental work has shown that 
exposure to high concentrations of NEFA can induce insulin resistance in 
insulin-sensitive tissues such as muscle and liver, and can impair 
pancreatic beta-cell production of insulin [@Rachek2014a;@Cnop2008a]. 
Observational and clinical studies have reported concordant findings, showing in
particular that elevated total plasma NEFA associates with an increased risk for
incident type 2 diabetes [@Paolisso1995a;@Salgin2012a].

Much of the previous experimental work on the role of NEFA in type 2 diabetes 
utilized individual fatty acids such as palmitic acid (16:0) or oleic acid 
(18:1n-9), or alternatively used specific oils such as soybean oil, which is 
high in the polyunsaturated fatty acid (PUFA) linoleic acid (18:2n-6), as the 
exposure to characterize the metabolic impact of total NEFA. However, fatty
acids comprise multiple molecules with diverse physiological functions, and few 
studies have analyzed the effects of a broader spectrum of fatty acids. For instance,
one study suggested that eicosapentaenoic acid (20:5n-3) can protect against the
lipotoxic effect of palmitic acid in the beta-cells [@Kato2008].

Despite a sizable literature studying the role of total NEFA concentration in 
diabetes, there are important gaps in this research. The majority of previous 
studies have used animal models or cell lines [@Maris2013a;@Deguil2011a], have
been short term human trials 
[@Szendroedi2012a;@Daniele2014a;@Liang2013a;@Kehlenbrink2012a], or have been
epidemiological studies that only looked at total NEFA and not individual fatty 
acids [@Miller2012a;@Johns2014a;@Charles1997a]. To date, there have been no
longitudinal studies examining the role of the composition of the serum NEFA
fraction on the pathogenesis of diabetes, which is critical given the protracted
natural history of diabetes and the growing appreciation of the divergent
effects of individual fatty acids. Therefore, our objective was to examine the
association of serum NEFA composition on differences over time in insulin
sensitivity and beta-cell function in a longitudinal cohort. We hypothesized
that higher palmitic acid and lower PUFA such as eicosapentaenoic acid in the
NEFA fraction would associate with declining insulin sensitivity and beta-cell
function over 6 years.

@@

Insulin sensitivity and beta-cell function indices were computed using the OGTT 
glucose and insulin data. Insulin sensitivity was assessed using HOMA2-%S [@Levy1998a]
and the Insulin Sensitivity Index (ISI) [@Matsuda1999]. HOMA2-%S largely
reflects hepatic insulin sensitivity, while ISI reflects whole-body insulin
sensitivity [@AbdulGhani2007].  Beta-cell function was assessed using the
Insulinogenic Index [@Wareham1995] over HOMA-IR (IGI/IR) and the Insulin
Secretion-Sensitivity Index-2 (ISSI-2) [@Retnakaran2009]. IGI/IR is a measure of
the first phase of insulin secretion while ISSI-2 is analogous to the
disposition index (but using OGTT values). Each index has been validated against
gold standard measures [@Hermans1999a;@Matsuda1999;@Retnakaran2009]. Formulas
used to calculate these measures can be found in the ESM Methods.

@@

All analyses were performed using R `r getRversion()` [@Rbase], along with the R
packages geepack `r packageVersion('geepack')` for GEE [@Hoejsgaard2006a]
and the package pls `r packageVersion('pls')` for the PLS analysis.
The R code and output for all analyses of this manuscript is available at 
https://dx.doi.org/10.6084/m9.figshare.3472433. Results were considered
statistically significance at p<0.05, after adjusting for multiple testing using
the Benjamini-Hochberg False Discovery Rate [@Benjamini1995a]. STROBE was used
as a guideline for reporting [@Vandenbroucke2007b].

@@

For this study, there were `r female$npct` females and `r european$npct` had
European-ancestry (see `r cite_t("basic")`). The mean follow-up time was 
`r calc_followup_time()` years, where `r calc_pct_full_visits()`% of 
participants attended all three visits.
The average age of the participants at baseline was `r aide::average(ds0yr$Age)`
years (`r aide::stddev(ds0yr$Age)` SD) and the average BMI was
`r aide::average(ds0yr$BMI)` (`r aide::stddev(ds0yr$BMI)` SD). Most of the 
participants, `r famhist$npct`, had a family history of diabetes. Insulin
sensitivity and beta-cell function measures showed a significant median
decline from baseline to 6-years of between `r chg_outcomes$chg` 
(p`r chg_outcomes$p`, based on a time-adjusted GEE model) in this analysis.
Consistent with this decline, there were `r calc_incid_dysgly("ConvertDM")`
and `r calc_incid_dysgly("ConvertPreDM")`incident cases of diabetes and
pre-diabetes (IFG and IGT), respectively, over the 6-years. Incident diabetes
cases were excluded from further GEE analyses. There were `r meds_use[1]` and
`r meds_use[2]` participants who used lipid and cholesterol modifying medications
at the baseline and 6-year visit, respectively. Medication use did not influence
NEFA composition, the outcomes, or the GEE models (data not shown).

@@

A number of associations were seen in the unadjusted GEE models 
(see `r cite_f('gee_unadj')`), particularly for NEFA modeled as a concentration.
There were several fatty acids (`r paste(unique(sig_fa_unadj$Xterms),collapse=", ")`)
that had significant associations with both insulin sensitivity and beta-cell
function outcomes. Most associations were negative, however, only 20:5n-3
(eicosapentaenoic acid; EPA) had positive associations with insulin sensitivity
as both a mol% and as a concentration. One fatty acid, 18:1n-7 (vaccenic acid),
had consistent negative associations with all four outcomes.

After full model adjustment (`r cite_f('gee_adj')`), most significant 
associations were attenuated, particularly with insulin sensitivity. The
remaining fatty acids (`r paste(unique(sig_fa_adj$Xterms),collapse=", ")`) had
significant associations, of which only EPA had a significant association (with
HOMA2-%S) as a mol% and was the only fatty acid with a positive association. All
other fatty acids had negative associations with IGI/IR and ISSI-2. The
magnitude of association for each of these variables was fairly consistent for
each beta-cell function measure. For every one SD increase in any of the NEFA
variables associated with beta-cell function, there was an average predicted
`r est_gee$est[1]`% lower IGI/IR and `r est_gee$est[2]`% lower ISSI-2 at any
given clinic visit.

@@

In a Canadian population of adults who were at-risk for diabetes, we found that 
higher total NEFA concentrations independently predicted lower beta-cell
function over 6 years. While we found negative associations with palmitic acid
(16:0), oleic acid (18:1n-9), and linoleic acid (18:2n-6) for the concentration
(nmol/mL) modeling, no associations were seen for these fatty acids as a
proportion, an observation that was further confirmed by multivariate cluster
techniques, which found no ability of these individual NEFA to predict beta-cell
function. We did find that one fatty acid as a mol%, eicosapentaenoic acid
(20:5n-3), had a positive association with insulin sensitivity. In general,
these observations suggest that the absolute size of the total NEFA fraction,
rather than its specific composition (aside from EPA), likely influences the 
pathogenesis of diabetes, at least within a population at-risk for diabetes.

@@

The role of total NEFA in the etiology of diabetes is well-documented.
Epidemiological studies have shown that higher NEFA associate with lower insulin
secretion and a higher risk for developing diabetes 
[@Paolisso1995a;@Salgin2012a;@Pankow2004a;@Charles1997a]. In a cross-sectional analysis of the
RISCK cohort, total NEFA had a negative association with insulin sensitivity and
a particularly strong negative association with beta-cell function
[@Johns2014a]. Experimentally, several potential mechanisms have been elucidated
for the role of NEFA on beta-cell function, particularly for palmitic acid.
Prolonged exposure to elevated NEFA can induce apoptosis in the beta-cells,
possibly through endoplasmic reticulum stress, formation of ceramides, and
generation of nitric oxide, as well as impairment of proinsulin production and
mitochondrial function [@Cnop2008a; @Maris2013a; @Giacca2011a; @Xiao2009a]. The
present analysis is the first study to our knowledge to examine the longitudinal
association of a broad spectrum of individual NEFA on beta-cell function in a
large cohort. We found that there was a strong signal of higher total NEFA,
palmitic acid, oleic acid, and linoleic acid modeled as concentrations with
lower beta-cell function. No fatty acid as a mol% had associations with
beta-cell function. Taken together, these results suggest that it is the
absolute size of the circulating NEFA fraction, irrespective of any specific 
composition of fatty acids, that is responsible for the hypothesized lipotoxic 
effects of NEFA on the beta-cells.

Biologically, in free-living populations chronic elevation of NEFA may be 
mediating its association with metabolic outcomes through TAG. In normal 
metabolism, NEFA enters the liver and assists in the production of TAG that is
to be processed into very-low density lipoproteins (VLDL) [@Nielsen2012a]. As
such, higher NEFA may contribute to hypertriglyceridemia, which is a known risk
factor for diabetes. In the sensitivity analysis adjusting for TAG, all
associations with beta-cell function were attenuated, suggesting that NEFA may
in fact be mediating its association with beta-cell dysfunction through higher
TAG.

@@

In conclusion, we found that total NEFA was a strong predictor for lower 
beta-cell function over 6 years, irrespective of the specific composition of the
NEFA fraction, suggesting that efforts at lowering circulating total NEFA 
through medication and/or lifestyle strategies to reduce the risk of diabetes 
are well warranted. We also found that higher EPA was associated with higher
hepatic insulin sensitivity. While future studies are needed to confirm these
findings, our results reinforce the importance of continuing to investigate the
role of circulating NEFA concentration on the natural history of diabetes.

>>>>

Title: Association of non-esterified fatty acid composition with insulin
sensitivity and beta cell dysfunction in the Prospective Metabolism and Islet Cell
Evaluation (PROMISE) cohort

@@

Aim: Our aim was to determine the longitudinal associations of individual non-esterified fatty acids (NEFA) on differences in insulin sensitivity (IS) and beta cell function over 6 years in a cohort of individuals who are at-risk for diabetes.

Methods: In the PROMISE cohort, 477 participants had serum NEFA (mol% and concentration) measured at the baseline visit and completed an OGTT at 3 time points over 6 years. Outcome variables were calculated using the OGTT values. IS was assessed using HOMA2-%S and the Matsuda index, while beta cell function was  assessed using the Insulinogenic index over HOMA-IR (IGI/IR) and the Insulin  Secretion-Sensitivity Index-2 (ISSI-2). Generalized estimating equations were  used, adjusting for time, waist, sex, ethnicity, baseline age, alanine  aminotransferase, and physical activity. NEFA were analyzed as both  concentrations (nmol/mL) and proportions (mol%) of the total fraction.

Results: Participants (73% female, 70% with European-ancestry) had their IS and beta cell function decline by 14-27% over 6 years of follow-up. While in unadjusted models there were several NEFA (e.g. 18:1n-7, 20:1n-9, 22:4n-6) that had negative associations with insulin sensitivity, nearly all of these associations were attenuated in fully adjusted models. In adjusted models, total NEFA, 16:0, 18:1n-9, and 18:2n-6 (as concentrations) had significant negative associations with IGI/IR and ISSI-2, while only 20:5n-3 (as mol%) had positive associations with HOMA2-%S.

Conclusions: Total NEFA concentration was a strong predictor of lower beta cell  function over 6 years. Our results suggest that the association with beta cell  function is due to the absolute size of the serum NEFA fraction, rather than the specific fatty acid composition.

@@

insulin sensitivity; beta cell function; longitudinal cohort; non-esterified
fatty acids; diabetes pathogenesis; fatty acid composition

@@

tab_basic <- tab('basic', 'Basic characteristics of the PROMISE participants at each of the three clinic visits.')
fig_dist <- fig('nefa', 'Concentrations (nmol/mL) of each NEFA in PROMISE participants at the baseline visit (2004-2006).')
fig_gee_unadj <- fig('gee_unadj', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta cell function over 6 years in the PROMISE cohort. Models are only adjusted for time. X-axis values represent a percent difference in the outcome per SD increase in the fatty acid. P-values were adjusted for the BH false discovery rate, with each increase in dot size and blackness representing a p-value significance of p>0.05, p<0.05, p<0.01, and p<0.001.')
fig_gee_adj <- fig('gee_adj', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta cell function over the 6 years in the PROMISE cohort. Generalized estimating equation models were fully adjusted. X-axis values represent a percent difference in the outcome per SD increase in the fatty acid. P-values were adjusted for the BH false discovery rate, with each increase in dot size and blackness representing a p-value significance of p>0.05, p<0.05, and p<0.01.')
```

@@

Total non-esterified fatty acids (NEFA) have been well documented to influence 
the pathogenesis of type 2 diabetes mellitus. Experimental work has shown that 
exposure to high concentrations of NEFA can induce insulin resistance in 
insulin-sensitive tissues such as muscle and liver, and can impair 
pancreatic beta cell production of insulin [@Rachek2014a;@Cnop2008a]. 
Observational and clinical studies have reported concordant findings, showing in
particular that elevated total plasma NEFA associates with an increased risk for
incident type 2 diabetes [@Paolisso1995a;@Salgin2012a].

Much of the previous experimental work on the role of NEFA in type 2 diabetes 
utilized individual fatty acids such as palmitic acid (16:0) or oleic acid 
(18:1n-9), or alternatively used specific oils such as soybean oil, which is 
high in the polyunsaturated fatty acid (PUFA) linoleic acid (18:2n-6), as the 
exposure to characterize the metabolic impact of total NEFA. However, fatty
acids comprise multiple molecules with diverse physiological functions, and few 
studies have analyzed the effects of a broader spectrum of fatty acids. For instance,
one study suggested that eicosapentaenoic acid (20:5n-3) can protect against the
lipotoxic effect of palmitic acid in the beta cells [@Kato2008].

Despite a sizable literature studying the role of total NEFA concentration in 
diabetes, there are important gaps in this research. The majority of previous 
studies have used animal models or cell lines [@Maris2013a;@Deguil2011a], have
been short term human trials 
[@Szendroedi2012a;@Daniele2014a;@Liang2013a;@Kehlenbrink2012a], or have been
epidemiological studies that only looked at total NEFA and not individual fatty 
acids [@Miller2012a;@Johns2014a;@Charles1997a]. To date, there have been no
longitudinal studies examining the role of the composition of the serum NEFA
fraction on the pathogenesis of diabetes, which is critical given the protracted
natural history of diabetes and the growing appreciation of the divergent
effects of individual fatty acids. Therefore, our objective was to examine the
association of serum NEFA composition on differences over time in insulin
sensitivity and beta cell function in a longitudinal cohort. We hypothesized
that higher palmitic acid and lower PUFA such as eicosapentaenoic acid in the
NEFA fraction would associate with declining insulin sensitivity and beta cell
function over 6 years.

@@

Insulin sensitivity and beta cell function indices were computed using the OGTT 
glucose and insulin data. Insulin sensitivity was assessed using HOMA2-%S [@Levy1998a]
and the Insulin Sensitivity Index (ISI) [@Matsuda1999]. HOMA2-%S largely
reflects hepatic insulin sensitivity, while ISI reflects whole-body insulin
sensitivity [@AbdulGhani2007].  beta cell function was assessed using the
Insulinogenic Index [@Wareham1995] over HOMA-IR (IGI/IR) and the Insulin
Secretion-Sensitivity Index-2 (ISSI-2) [@Retnakaran2009]. IGI/IR is a measure of
the first phase of insulin secretion while ISSI-2 is analogous to the
disposition index (but using OGTT values). Each index has been validated against
gold standard measures [@Hermans1999a;@Matsuda1999;@Retnakaran2009]. Formulas
used to calculate these measures can be found in the ESM Methods.

@@

All analyses were performed using R `r getRversion()` [@Rbase], along with the R
packages geepack `r packageVersion('geepack')` for GEE [@Hoejsgaard2006a]
and the package pls `r packageVersion('pls')` for the PLS analysis.
The R code and output for all analyses of this manuscript is available at 
https://dx.doi.org/10.6084/m9.figshare.3472433. Results were considered
statistically significance at p<0.05, after adjusting for multiple testing using
the Benjamini-Hochberg False Discovery Rate [@Benjamini1995a]. STROBE was used
as a guideline for reporting [@Vandenbroucke2007b].
The cohort datasets analysed during the current study are available from the
corresponding author on reasonable request.

@@

For this study, there were `r female$npct` females and `r european$npct` had
European-ancestry (see `r cite_t("basic")`). The mean follow-up time was 
`r calc_followup_time()` years, where `r calc_pct_full_visits()`% of 
participants attended all three visits.
The average age of the participants at baseline was `r aide::average(ds0yr$Age)`
years (`r aide::stddev(ds0yr$Age)` SD) and the average BMI was
`r aide::average(ds0yr$BMI)` (`r aide::stddev(ds0yr$BMI)` SD). Most of the 
participants, `r famhist$npct`, had a family history of diabetes. Insulin
sensitivity and beta cell function measures showed a significant median
decline from baseline to 6-years of between `r chg_outcomes$chg` 
(p`r chg_outcomes$p`, based on a time-adjusted GEE model) in this analysis.
Consistent with this decline, there were `r calc_incid_dysgly("ConvertDM")`
and `r calc_incid_dysgly("ConvertPreDM")`incident cases of diabetes and
pre-diabetes (IFG and IGT), respectively, over the 6-years. Incident diabetes
cases were excluded from further GEE analyses. There were `r meds_use[1]` and
`r meds_use[2]` participants who used lipid and cholesterol modifying medications
at the baseline and 6-year visit, respectively. Medication use did not influence
NEFA composition, the outcomes, or the GEE models (data not shown).

@@

A number of associations were seen in the unadjusted GEE models 
(see `r cite_f('gee_unadj')`), particularly for NEFA modeled as a concentration.
There were several fatty acids (`r paste(unique(sig_fa_unadj$Xterms),collapse=", ")`)
that had significant associations with both insulin sensitivity and beta cell
function outcomes. Most associations were negative, however, only 20:5n-3
(eicosapentaenoic acid; EPA) had positive associations with insulin sensitivity
as both a mol% and as a concentration. One fatty acid, 18:1n-7 (vaccenic acid),
had consistent negative associations with all four outcomes.

After full model adjustment (`r cite_f('gee_adj')`), most significant 
associations were attenuated, particularly with insulin sensitivity. The
remaining fatty acids (`r paste(unique(sig_fa_adj$Xterms),collapse=", ")`) had
significant associations, of which only EPA had a significant association (with
HOMA2-%S) as a mol% and was the only fatty acid with a positive association. All
other fatty acids had negative associations with IGI/IR and ISSI-2. The
magnitude of association for each of these variables was fairly consistent for
each beta cell function measure. For every one SD increase in any of the NEFA
variables associated with beta cell function, there was an average predicted
`r est_gee$est[1]`% lower IGI/IR and `r est_gee$est[2]`% lower ISSI-2 at any
given clinic visit.

@@

In a Canadian population of adults who were at-risk for diabetes, we found that 
higher total NEFA concentrations independently predicted lower beta cell
function over 6 years. While we found negative associations with palmitic acid
(16:0), oleic acid (18:1n-9), and linoleic acid (18:2n-6) for the concentration
(nmol/mL) modeling, no associations were seen for these fatty acids as a
proportion, an observation that was further confirmed by multivariate cluster
techniques, which found no ability of these individual NEFA to predict beta cell
function. We did find that one fatty acid as a mol%, eicosapentaenoic acid
(20:5n-3), had a positive association with insulin sensitivity. In general,
these observations suggest that the absolute size of the total NEFA fraction,
rather than its specific composition (aside from EPA), likely influences the 
pathogenesis of diabetes, at least within a population at-risk for diabetes.

@@

The role of total NEFA in the etiology of diabetes is well-documented.
Epidemiological studies have shown that higher NEFA associate with lower insulin
secretion and a higher risk for developing diabetes 
[@Paolisso1995a;@Salgin2012a;@Pankow2004a;@Charles1997a]. In a cross-sectional analysis of the
RISCK cohort, total NEFA had a negative association with insulin sensitivity and
a particularly strong negative association with beta cell function
[@Johns2014a]. Experimentally, several potential mechanisms have been elucidated
for the role of NEFA on beta cell function, particularly for palmitic acid.
Prolonged exposure to elevated NEFA can induce apoptosis in the beta cells,
possibly through endoplasmic reticulum stress, formation of ceramides, and
generation of nitric oxide, as well as impairment of proinsulin production and
mitochondrial function [@Cnop2008a; @Maris2013a; @Giacca2011a; @Xiao2009a]. The
present analysis is the first study to our knowledge to examine the longitudinal
association of a broad spectrum of individual NEFA on beta cell function in a
large cohort. We found that there was a strong signal of higher total NEFA,
palmitic acid, oleic acid, and linoleic acid modeled as concentrations with
lower beta cell function. No fatty acid as a mol% had associations with
beta cell function. Taken together, these results suggest that it is the
absolute size of the circulating NEFA fraction, irrespective of any specific 
composition of fatty acids, that is responsible for the hypothesized lipotoxic 
effects of NEFA on the beta cells.

Biologically, in free-living populations chronic elevation of NEFA may be 
mediating its association with metabolic outcomes through TAG. In normal 
metabolism, NEFA enters the liver and assists in the production of TAG that is
to be processed into very-low density lipoproteins (VLDL) [@Nielsen2012a]. As
such, higher NEFA may contribute to hypertriglyceridemia, which is a known risk
factor for diabetes. In the sensitivity analysis adjusting for TAG, all
associations with beta cell function were attenuated, suggesting that NEFA may
in fact be mediating its association with beta cell dysfunction through higher
TAG.

@@

In conclusion, we found that total NEFA was a strong predictor for lower 
beta cell function over 6 years, irrespective of the specific composition of the
NEFA fraction, suggesting that efforts at lowering circulating total NEFA 
through medication and/or lifestyle strategies to reduce the risk of diabetes 
are well warranted. We also found that higher EPA was associated with higher
hepatic insulin sensitivity. While future studies are needed to confirm these
findings, our results reinforce the importance of continuing to investigate the
role of circulating NEFA concentration on the natural history of diabetes.
